---
title: "The effectiveness of cultural inheritance as a problem solving strategy"
author:
  - { name: Pierce Edmiston, affiliation_id: 1 }
  - { name: Gary Lupyan, affiliation_id: 1 }
  - { name: Maxime Derex, affiliation_id: 2 }
affiliation:
  - id: 1
    department: Department of Psychology
    university: University of Wisconsin-Madison
  - id: 2
    department: Biosciences Department
    university: University of Exeter
email: pedmiston\@wisc.edu
address: 1202 W. Johnson St., Department of Psychology, University of Wisconsin, Madison, WI 53703.
bibliography: "references.bib"
csl: "template/plos-one.csl"
output:
  bookdown::pdf_document2:
    template: template/plos-one.tmpl
    keep_tex: yes
    number_sections: no
abstract: |
  Human cultures are able to accumulate bodies of problem solving knowledge that
  no individual could create on their own. The cultural transmission of problem
  solving knowledge involves both vertical transmission from one generation to
  the next, and horizontal transmission among peers in the same generation.
  Although both forms of cultural transmission have been studied independently,
  they have yet to be compared directly. We designed an experiment to tease
  apart the effectiveness of vertical and horizontal transmission relative to
  individuals solving the same set of problems alone. We found that vertical
  transmission of problem solving knowledge resulted in reliable improvements in
  performance over generations. The amount of improvement was similar to what
  could be achieved via horizontal transmission, and by giving individuals twice
  as long to work on the task. Our results are intended to improve our
  understanding of cumulative cultural evolution, primarily the mechanisms by
  which groups can arrive at solutions to problems that individuals cannot. We
  believe this study establishes an experimental paradigm for studying the
  effectiveness of cultural inheritance as a problem solving strategy in
  other problem solving domains.
---
```{r team-structures-config, include=FALSE}
library(knitr)
library(crotchet)

opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = "hide",
  fig.width = 3,
  fig.height = 3,
  fig.crop = F
)

# Read all knitr chunks in external .R file
read_chunk("R/team-structures.R")
read_chunk("R/types-of-time.R")
```
```{r team-structures}
```
```{r types-of-time}
```

# Introduction {-}

Humans are not the only tool users in the animal kingdom [@Beck:1980animal].
What sets human tool users apart may not be our ability to make tools, but our
ability to use and improve tools that were first created by other people. This
ability is often referred to as the "ratchet effect" [@Tomasello:1993vh;
@Dean:2013dp], and is believed to be an important component of cumulative
cultural evolution. The clearest evidence of cumulative cultural evolution is
the observation that cultures, through generations of incremental improvement,
can sustain complex tools and other solutions to problems that no individual
could recreate on their own [@Boyd:2011gp]. The purpose of this research was to
better understand the effectiveness of cultural transmission in accumulating
solutions to problems that no individual could recreate on their own.

Groups of problem solvers can share information via both vertical and
horizontal forms of cultural transmission [@Boyd:1985cu]. Vertical transmission
involves passing cultural products from parent to child, whereas in horizontal
transmission, cultural products are passed among peers in the same generation.
Although both forms of cultural transmission involve social learning---learning
that is enabled by the observation of or interaction with others
[@Rendell:2011gk]---the consequences of either vertical or horizontal
transmission on culture accumulation have not been directly compared. Here we
tested whether problem solving knowledge that is inherited at the start of
a problem solving task via vertical transmission has a different impact on
problem solving performance than purportedly the same knowledge shared
horizontally among problem solvers working in parallel.

Previous research on the differences between vertical and horizontal
transmission has found that some cultural products, such as religious beliefs,
are more likely to be learned from parents than from peers [@Newcomb:1937vy;
@Boyd:1985cu]. Whether individuals learn from parents or peers may also
have consequences for the distribution of traits in a population. For example,
@Bettinger:1999vv surveyed projectile points recovered in the Great Basin
region, and found that in one region, the point designs were highly correlated
with one another, whereas in a different region, the point designs were poorly
correlated. They argued that these differences could be attributed to different
forms of cultural transmission. Highly similar points are the expected result
of a horizontal transmission process where members of the society copy the most
prestigious hunters, whereas uncorrelated points would result from vertical
transmission where each family innovated their own designs. This interpretation
has been supported by experimental and computational simulations showing that
groups of problem solvers result in greater convergence and smaller variation
in design than than independently evolving lineages [@Mesoudi:2008vi;
@Mesoudi:2008we].

But in explaining how groups of problem solvers can solve problems that
individuals cannot, previous research has focused on horizontal transmission
while ignoring the potential role of vertical transmission. For example,
@Derex:2015cbb had groups and individuals complete the same problem solving
task, and found that a group strategy involving horizontal transmission
proved more effective for discovering complex solutions than an isolated
strategy. The adaptiveness of horizontal transmission over individual problem
solving has been attributed to an advantage of parallel search in that groups
are better able to search the space of possible solutions than individuals
working alone [@Acerbi:2016ej]. Horizontal transmission among groups of
problem solvers has been shown to be adaptive for a wider range of problem
solving scenarios than individual learning [@Golman:2010fz; @Rendell:2010go].

When vertical transmission has been studied directly, it has often not been
compared to alternative strategies for solving the same problems. For example,
@Caldwell:2008gf had participants create paper airplanes or spaghetti towers in
a transmission chain design involving vertical transmission, where new
generations of problem solvers learned from the previous generation. The
results showed that performance tended to improve along the transmission
chains, suggesting that human problem solvers are skilled at using and
improving solutions to problems that were created by others. But since the
performance of the transmission chains was not compared to performance of
isolated individuals given the same opportunities, we do not know whether
vertical transmission by itself can result in the accumulation of problem
solving knowledge that no individual could recreate on their own.

To better understand the effectiveness of vertical transmission as a strategy
for solving problems, we conducted an experiment where participants completed
the same problem solving task using one of three strategies. We refer to the
strategy involving vertical transmission as **diachronic collaboration**. In
diachronic collaboration, participants completed the problem solving task one
at a time, and at the end of the session, they passed the solutions they had
discovered on to the next generation. The next generation, rather than starting
from scratch, starts off with the solutions inherited from the previous
generation. In synchronic collaboration, participants work in a team on the
same problem solving task at the same time and in parallel, sharing solutions
with each other as they are discovered. Both diachronic and synchronic
strategies are compared to an isolated strategy in which individuals work alone
on the same problems.

Part of the reason vertical and horizontal cultural transmission have not been
compared before is because they are believed to involve similar learning
mechanisms. This makes the distinction between vertical and horizontal cultural
transmission different from the analogous processes in genetic evolution. In
genetic evolution, vertical transmission occurs via reproduction, whereas
horizontal transmission is the transfer of genetic material by any means _other_
than reproduction [@Soucy:2015gp]. Thus, the difference between vertical and
horizontal transmission of genetic material is motivated by an underlying
difference in mechanism, as the cellular mechanisms involved in reproduction
(e.g., meiosis) are different from those involved in horizontal transfer (e.g.,
conjugation). In contrast, there is no such difference in mechanism between the
vertical and horizontal transmission of cultural products. All processes of
cultural transmission involve social learning, in which one person learns a
behavior from the observation or interaction with someone else. The cognitive
mechanisms that allow for social learning are functionally the same regardless
of who is being copied.

At the same time, the process of learning from the accumulated knowledge of the
previous generation is on the surface much different than accumulating the same
knowledge in parallel as a part of a team. Accurately measuring the benefits
of cultural inheritance requires taking the benefits to inheritance relative
to the cost of creating it. This control is necessary because we cannot complete
the within-subject design where individuals solve the problem individually
from nothing and after inheriting from someone else.

In our comparison of diachronic, synchronic, and isolated strategies, we
controlled for labor time. Labor time can be contrasted with calendar time
and learning time (Fig. \@ref(fig:types-of-time)). We controlled for labor time
so that our results can be interpreted in terms of the effect each problem
solving strategy has on individual problem solving ability. Without controlling
for labor time, any differences between the three strategies are more difficult
to interpret. For example, if only calendar hours are controlled, an isolated
individual given 30 minutes could theoretically be compared to a synchronic
team given 30 minutes _regardless of the size of the synchronic team_. This
fact makes it difficult to understand the effect of teamwork on individual
performance when only controlling for calendar time [@Propp:2003kj;
@Tindale:1992ub].

(ref:types-of-time) Team structures controlling for labor time. **Left.** Team
structures. Each node is a person. **Middle.** Relationship between labor time
and calendar time. Synchronic teams spend labor time more quickly than the
other strategies. **Right.** Learning times by strategy, showing that an
isolated strategy provides the most time for individual learning.

```{r types-of-time, fig.width=5, fig.height=2, fig.cap="(ref:types-of-time)"}
gridExtra::grid.arrange(
  crotchet::read_graphviz("team-structures", package = "totems"),
  gg_time,
  gg_person,
  nrow = 1,
  widths = c(0.25, 0.375, 0.375)
)
```

Our experiment measured the effectiveness of the simplest form of diachronic
inheritance involving two generations of problem solvers collaborating via
vertical transmission.

# Methods {-}

## Team structures {-}

Participants were randomly assigned to complete the same problem solving task in
different team structure conditions (Fig. \@ref(fig:team-structures-50min)).
Participants in the diachronic condition worked on the problem solving task one
at a time. At the end of a diachronic participant's session, the solutions they
discovered were passed on to be inherited by a single participant assigned to
the next generation of the same team. Participants assigned to the synchronic
team structure worked independently and in parallel in teams of two. Their labor
was coordinated such that solutions discovered by one team member were
immediately shared with the other members of the team. Isolated participants
completed the problem solving task as individuals, only benefiting from the
solutions they could discover on their own. Isolated participants completed the
experiment in a single 50 minute session.

(ref:team-structures-50min) Team structures.

```{r team-structures-50min, fig.cap="(ref:team-structures-50min)"}
crotchet::draw_graphviz("team-structures-50min", package = "totems")
```

## Problem solving task {-}

Participants played the Totem game adapted from @Derex:2015cbb. Their task was
to discover tools through the combinations of items that could be used to create
"a sacred totem to appease the gods." To build a totem, participants first
needed to construct an axe out of three independently discovered tools: a
refined stick used as a handle, a sharpened rock for the blade, and a string
wound from bark fibers for binding (Fig. \@ref(fig:landscape-sample)). More
advanced tools produce larger and more intricate totems.

(ref:landscape-sample) A sample of the solution landscape. The top row of 6
items were available to problem solvers at the start of the game. New tools
could be produced through the combination of different items (more than one
arrow points to the item) or through the refinement of a single item (a single
arrow points to the item). The axe is required to construct the first totem
pole.

```{r landscape-sample, fig.width=5, fig.cap="(ref:landscape-sample)", eval=FALSE}
draw_image("landscape-sample.gv", package = "totems")
```

Once a tool was discovered, the recipe for its creation---a list of the items
that had to be combined in order to create the tool---was recorded in an
innovation record. Participants could review their past discoveries by clicking
on a tool to see its recipe.

Innovation records were also used as the means for sharing solutions among
members of a team and across sessions. Diachronic participants inherited an
innovation record at the start of the experiment. Participants
assigned to the synchronic condition could view the innovation records of each
of their teammates at any point during the session. Once someone on a team
discovered a new tool, the recipe for its creation could be viewed by all members of
the team.

(ref:interface) The Totem game. Participants generated guesses
by dragging items into the **Workshop** and selecting the "Try" button. If the
guess created a tool, it could be dragged into the **Stock** panel, and used
again in later combinations. Once segments of a totem were made through
cutting, carving, and painting, they could be dragged into the **Totem** panel
for scoring.

```{r interface, fig.width=6, fig.cap="(ref:interface)"}
draw_image("gameplay", "totems")
```

## Participants {-}

Participants were undergraduate students at the University of Wisconsin-Madison
who completed the experiment in exchange for course credit. Students received
credit based on the duration of each completed session. Sessions were either 25
minutes or 50 minutes. Participant and team recruitment is presented in Table
\@ref(tab:counts-table).

(ref:condition-counts) Participant and team recruitment.

```{r counts-table, results="asis"}
kable(N, caption = "(ref:condition-counts)")
```

## Data {-}

The data were collected as a part of a larger research project using the
Totem game. The results of all experiments using this paradigm
are openly available on GitHub[^github] and the Open Science Framework[^osf].

[^github]: https://github.com/pedmiston/totems-data/
[^osf]: https://osf.io/b8c3y/

# Results {-}

We compared the performance of participants assigned to use the diachronic,
synchronic, and isolated strategies in three ways. First we compared performance
based on the total number of tools that were discovered. Second we report
guessing efficiency on a per-tool basis. Finally, we compared the guessing
strategies used by participants in each condition.

## Total tools {-}

The total number of tools discovered using each of the three strategies is
shown in Fig. \@ref(fig:innovations-50min). Our primary hypothesis concerned
the effectiveness of a diachronic strategy as compared to the alternative
strategies for solving the same problems in the same amount of time. To test
this, we fit a hierarchical linear model predicting the total number of tools
discovered by each participant based on strategy, where strategy was
dummy-coded such that second generation diachronic participants were the
reference group. The hierarchical model included random intercepts for each
team. The results showed that a diachronic strategy was similarly effective to
either alternative strategy. Second generation diachronic participants
discovered more tools on average than isolated individuals, but this
difference may be due to chance, `r results$DG2_v_I50`. Conversely,
diachronic teams discovered fewer tools on average than two-person
synchronic teams, but again this difference may not be reliable,
`r results$S2_v_DG2`.

We also tested the effectiveness of teamwork compared to solving the same
problems as individuals. To test this, we compared two-person teams (both
diachronic and synchronic) to isolated individuals. In the hierarchical linear
model, team strategy was coded with Helmert contrasts, and intercepts
for team were included as random effects. Teams discovered more tools than individuals,
`r results$teamwork_stats`, without a significant residual difference between the
diachronic and synchronic strategies, `r results$teamwork_residual`. Post hoc
tests revealed that this effect of teamwork was driven primarily by synchronic
teams discovering more tools than isolated individuals, `r results$S2_v_I50`.

(ref:innovations-50min) Tools discovered in 50 labor minutes by strategy.
**A.** Rate of innovation by labor hours. Each line is the average number of
tools discovered by participants in each condition. **B.** Total
tools by strategy. Each point is a single participant. Bars with errors
show model predictions with ±1 standard error.

```{r innovations-50min, fig.width=5, fig.height=3, fig.cap="(ref:innovations-50min)"}
grid.arrange(
  innovation_rate_50min_plot + ggtitle("A"),
  num_innovations_50min_plot + ggtitle("B"),
  nrow = 1,
  widths = c(0.55, 0.45)
)
```

## Guesses per tool {-}

In addition to comparing the three strategies in terms of total number of
tools discovered, we also performed a more fine-grained analysis of guessing
behaviors. We compared the three strategies in terms of how many guesses were
required to discover each tool. To count the number of guesses that were
required for each tool, we tallied all guesses made from the moment in which
a new tool was eligible for discovery until that tool was discovered. A new
tool was eligible for discovery once all of the items required to create the
new tool had been discovered. We refer to new tools that are eligible for
discovery as "adjacent", since they are adjacent to the current inventory in
the solution landscape. The minimum number of adjacent tools for any inventory
was one, indicating that there was a single tool that must be discovered before
new tools can be created. The median number of adjacent tools was four. When
there was more than one adjacent tool, each incorrect guess was counted toward
the total number of guesses required for each of the adjacent tools.

As expected, we found the second generation diachronic participants benefited
from inheritance in that they required significantly fewer guesses to produce
each tool that they inherited relative to the same tools being discovered
by participants in conditions without inheritance,
`r results$guesses_per_item_inheritance` (Fig. \@ref(fig:guesses-per-innovation)A).

However, this benefit to inheritance in terms of fewer guesses per
tool did not extend to tools that had not been discovered by the previous
generation. Comparing the number of guesses per tool for only the tools that
were newly discovered by the second generation diachronic participants did not
reveal a benefit to inheritance that extended beyond the inherited solutions,
`r results$guesses_per_new_item_inheritance`
(Fig. \@ref(fig:guesses-per-innovation)B).

(ref:guesses-per-innovation) Guesses per tool by inheritance condition. Each
thin gray line shows the average number of guesses for a particular tool. The
"No inheritance" condition includes guesses from synchronic and isolated
participants, as well as first generation diachronic players. The inheritance
condition includes second generation diachronic participants only. **A.**
Guesses per tool for tools inherited from an ancestor. **B.** Guesses per tool
for new tools, not inherited from an ancestor.

```{r guesses-per-innovation, fig.width=5, fig.cap="(ref:guesses-per-innovation)"}
gridExtra::grid.arrange(
  guesses_per_item_plot + ggtitle("A"),
  guesses_per_new_item_plot +
    ggtitle("B") +
    theme(
      axis.title.y = element_blank(),
      axis.text.y = element_blank()
    ),
  nrow = 1,
  widths = c(0.55, 0.45)
)
```

## Guessing strategy {-}

The analyses of guessing effectiveness reported above treat all incorrect
guesses as the same, but in reality some guesses are more informative than
others. Specifically, an optimal guessing strategy is to only attempt unique
guesses that have never been tried before. However, participants often attempt
redundant guesses, thus employing a suboptimal guessing strategy. To quantify
the guessing strategies used by participants in each team strategy, we
classified each guess into one of four types, listed in Table
\@ref(tab:guess-types), and tested whether team strategy had an impact on
guessing strategy.

(ref:guess-types) Guess types. Each guess can be categorized into one of four
types. Redundant guesses were incorrect guesses made at least once before.
Repeat items were guesses that recreated previously discovered tools.
Unique guesses were incorrect guesses never made. Unique items were unique
guesses that created a new tool. Guess types were classified within-subject.

```{r guess-types, results="asis"}
recode_guess_type() %>%
  select(`Guess type` = GuessTypeLabel,
         `Unique guess` = UniqueGuess,
         `Created item` = CreatedItem,
         `Unique result` = UniqueResult) %>%
  kable(caption = "(ref:guess-types)")
```

The guessing strategies for participants in each team strategy are shown in Fig.
\@ref(fig:guess-props). For our analysis, we focused on the proportions of
redundant and unique guesses, corresponding to inefficient and efficient
guessing strategies, respectively.

Participants in all strategies generated a significant number of redundant
guesses. We found that second generation diachronic participants generated
a lower proportion of redundant guesses than
participants in the other strategies, `r results$prop_redundant_inheritance`.
Post hoc comparisons revealed that second generation diachronic participants
had a lower proportion of redundant guesses than first generation
diachronic participants, `r results$prop_redundant_dg2`, and isolated
individuals, `r results$prop_redundant_i50`, but not synchronic participants,
`r results$prop_redundant_i50`.

Turning to the unique guesses, we found no difference between
second generation diachronic participants and the other strategies,
`r results$prop_unique_inheritance`. Pairwise comparisons suggested
a potential difference between second generation diachronic
and isolated participants, `r results$prop_unique_i50`, but not between
second and first generation diachronic participants,
`r results$prop_unique_dg2`, and not between second generation diachronic
and synchronic participants, `r results$prop_unique_s2`.

(ref:guess-props) Guessing strategies by team structure. See Table
\@ref(tab:guess-types) for a description of the types of guesses. DG1 and DG2 refer
to first and second generation diachronic participants. I50 refers to isolated
participants given 50 minutes. S2 refers to two-person synchronic teams.

```{r guess-props, fig.width=5, fig.cap="(ref:guess-props)"}
prop_guess_types_50min_plot
```

# Discussion {-}

We sought to better understand the process by which groups of problem solvers
can arrive at solutions to problems that no individual could on their own.
Previous research addressing this question has found that groups of problem
solvers collaborating via horizontal transmission can in some cases raise the
performance of the group beyond what would be likely to be achieved
individually [@Derex:2015cbb]. We extended this line of research by studying
the effectiveness of vertical transmission, where individuals start off
a problem solving task by inheriting directly from a previous generation.

Our results suggest that inheriting from a previous generation at
the beginning of a problem solving task may have different consequences than
acquiring purportedly the same information by other means. The reason may be
related to the unique learning environment experienced in vertical
transmission. In vertical transmission, generations start off with
access to an accumulated body of problem solving knowledge. These
learning experiences may be different than acquiring purportedly the same
knowledge after initial learning has already taken place. As a result, vertical
transmission may be able to promote problem solving behaviors that could not be
achieved individually.

Although the purpose of the experiments reported here was to disentangle
the effects of vertical and horizontal transmission, future work must
address the more natural forms of collaboration involving both
diachronic and synchronic elements.

In seeking to explain how and why humans are skilled problem solvers, many have
pointed to our ability to inherit skilled behaviors learned by others through
cultural transmission. But whether cultural transmission has an impact on
future problem solving performance, beyond what was inherited, is an open area
of research. Here we establish that vertical transmission is a unique form
a problem solving worthy of further investigation.

# References {-}

\noindent
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{6pt}
