---
title: |
  The effects of cultural inheritance  
  on problem solving ability
author:
  \textbf{Pierce Edmiston}$^1$, Maxime Derex$^2$, Gary Lupyan$^1$
bibliography: references.bib
csl: apa.csl
output:
  beamer_presentation:
    theme: metropolis
    keep_tex: yes
    template: template/slides.tmpl
---
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(crotchet)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
read_chunk("team-structures.R")
```
```{r team-structures, include=FALSE}
```
```{r config, include=FALSE, cache=FALSE}
t_$base_theme$text$size <- 20
t_$annotation_size <- 5
theme_set(t_$base_theme)
```

# Cumulative cultural evolution

Populations solve increasingly complex problems over generations  
that individuals struggle to solve on their own.

```{r computer, fig.cap="The Difference Engine and the Apple Watch S3."}
gridExtra::grid.arrange(
  crotchet::read_image("img/difference-engine.jpg", width = 0.9),
  crotchet::read_image("img/apple-watch.jpg", width = 0.9),
  nrow = 1
)
```


Here are two computers.

The one on the left is a portion of the Difference Engine, a one-of-a-kind
machine designed by Charles Babbage in the 1820s, but famously never actually
completed, or at least not in his lifetime. The difference engine was a machine
designed to compute arbitrary functions, so it was a very early computer.

On the right, we have the inside of an Apple Watch S3 --- a solid state
computer, with no moving parts. Today, about 4 million of these watches
are created every quarter by Apple, a company worth over one trillion
dollars with over 120,000 employees.

By just about every measure I can think of, the Apple Watch is a much better
computer than Babbage's 15 ton machine, and this is the kind of cumulative
cultural evolution I think we can all agree on.

Human populations solve increasingly complex problems over generations
that individuals struggle to solve on their own.

The question I'm interested in is WHY.
Why are cultures better at solving problems than individuals?


# Why are cultures better at solving problems than individuals?

Inheritance
:   Knowledge can be passed to the next generation.

Group size
:   Individuals can coordinate labor in large groups.

**Does culture have an impact on problem solving ability?**


There are two reasons I want to talk about today. The first is the idea
of inheritance: that we can pass our knowledge from one generation to
the next, which means that problem solving can extend far longer than any
individual lifetime.

The second reason is related to group size. This is the idea that individuals
can coordinate their labor in groups to accomplish something that they couldn't
by themselves.

To me, as a psychologist, the most interesting question is whether
culture in either case actually has an impact on problem solving ability.
Does problem solving just add up as you add more generations or more
people to the group, or are there interactions, places where the group
performs better than the sum of its parts?

I'm going to argue that we actually know a fair amount about the effects
of group size on problem solving ability, but we actually don't know very much
about the effects on inheritance on problem solving.


# When does the group outperform the best individual?

**Not very often!**

- Search tasks[^search]
- Perceptual decision making[^decision]

[^search]: @Hong:2004gt; @Derex:2016hj
[^decision]: @Bahrami:2010jl


When does the group outperform the best individual?

From the perspective of cognitive psychology,
the answer to this question is "not very often". For most problem solving tasks
that individuals can do alone, making a group do the same task doesn't
usually make them do it any better.

But this makes the cases where groups do seem to benefit
individuals all the more interesting.

One class of problems where groups generally tend to do better
is in search tasks. These are cases where people need to search
a space of possible solutions looking for the best ones. In these
cases, working in a group tends to help by widening the search,
and making it easier to exploit optimal solutions.

Another class of problems is in perceptual decision making under uncertainty,
where communicating with someone else about what you think you saw can improve
the performance of the group beyond what would have been achieved by the
best individual working alone.


# When is it more effective to inherit from a previous generation?

**We don't know!**

- Previous experiments have been interested in _capacity_, not _efficiency_.
- Models have been used to argue that it doesn't need to be particularly efficient.


Now let me ask a parallel question about the effects of inheritance on
problem solving.

When is it more effective to inherit from a previous generation as opposed
to solving the same set of problems some other way?

I don't think we really have great answers to this question yet.
Transmission chain experiments have typically been interested
in the capacity to accumulate improvements, not the efficiency
of the process as a problem solving strategy.

Also, theory and models have been used to argue that cultural inheritance
doesn't need to be particularly efficient in order to be beneficial.

So how do we answer this question?


# Measuring the effects of cultural inheritance on problem solving

```{r team-structures-1, out.height="80%", fig.align="center"}
crotchet::draw_graphviz("team-structures", package = "totems")
```


Here is my answer to this question. If we want to understand the effects
of cultural inheritance on problem solving, we need to compare this strategy,
which I refer to as "diachronic problem solving", to alternative ways
of solving the same set of problems. The two alternative strategies I'm
most interested in are isolated individuals working alone, and synchronic
problem solving, which is a more traditional group work structure where
people are working in parallel.


# Types of time

```{r types-of-time, fig.width=5, fig.height=2, fig.cap="Team structures controlling for labor time."}
gridExtra::grid.arrange(
  crotchet::read_graphviz("team-structures", package = "totems"),
  gg_time,
  gg_person,
  nrow = 1,
  widths = c(0.25, 0.375, 0.375)
)
```




# All team structure conditions

```{r team-structures-all, out.height="90%", fig.align="center"}
crotchet::draw_graphviz("team-structures-all", package = "totems")
```




# The Totem Game [@Derex:2015cbb]

```{r gameplay, out.height="90%"}
draw_image("gameplay", package = "totems")
```




# Recipes

```{r recipes, out.height="80%"}
crotchet::draw_image("recipes.gv", package = "totems")
```




# Solution landscape

```{r landscape, out.height="80%"}
crotchet::draw_image("landscape-sample.gv", package = "totems")
```




# Experiment 1: Simple diachronic inheritance

```{r team-structures-50min, fig.align="center"}
crotchet::draw_graphviz("team-structures-50min", package = "totems")
```

# Experiment 1: Simple diachronic inheritance

```{r innovations-50min}
grid.arrange(
  innovation_rate_50min_plot + ggtitle("A"),
  num_innovations_50min_plot + ggtitle("B"),
  nrow = 1,
  widths = c(0.55, 0.45)
)
```

# Experiment 2: Diachronic versus individual inheritance

```{r team-structures-selfother}
crotchet::draw_graphviz("team-structures-selfother", package = "totems")
```

# Experiment 2

```{r innovations-by-generation-selfother}
innovations_by_generation_plot
```

# Experiment 2

```{r diachronic-v-isolated-inheritance}
gridExtra::grid.arrange(
  learning_times_plot + ggtitle("A"),
  first_discovery_by_generation_plot + ggtitle("B"),
  ncol = 1
)
```

# Experiment 3

```{r scalability}
max_innovations_by_teamsize_plot
```

# Experiment 3

```{r scalability-simulations}
gridExtra::grid.arrange(
  max_innovations_by_teamsize_plot,
  bots_team_size_plot,
  nrow = 1
)
```

# Summary

Can cultural inheritance improve the problem solving abilities of future generations?

- A diachronic strategy is comparable when controlling for labor time.
- A diachronic strategy remediates the cost of fixation.
- A diachronic strategy is more scalable than a sychronic strategy.

# The effects of cultural inheritance on problem solving ability

**Pierce Edmiston**, Max Derex, Gary Lupyan  
pedmiston@wisc.edu  
[github.com/pedmiston/team-structures](https://github.com/pedmiston/team-structures)

# References {.allowframebreaks}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.25in}
\noindent
\footnotesize
