---
title: |
  The effects of cultural inheritance  
  on problem solving ability
author:
  \textbf{Pierce Edmiston}$^1$, Maxime Derex$^2$, Gary Lupyan$^1$
bibliography: references.bib
csl: apa.csl
output:
  beamer_presentation:
    theme: metropolis
    keep_tex: yes
    template: template/slides.tmpl
---
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(crotchet)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
read_chunk("team-structures.R")
```
```{r team-structures, include=FALSE}
```
```{r config, include=FALSE, cache=FALSE}
t_$base_theme$text$size <- 20
t_$annotation_size <- 5
theme_set(t_$base_theme)
```

# Cumulative cultural evolution

Populations solve increasingly complex problems over generations  
that individuals struggle to solve on their own.

```{r computer, fig.cap="The Difference Engine and the Apple Watch S3."}
gridExtra::grid.arrange(
  crotchet::read_image("img/difference-engine.jpg", width = 0.9),
  crotchet::read_image("img/apple-watch.jpg", width = 0.9),
  nrow = 1
)
```

<!--
Here are two computers.

The one on the left is a portion of the Difference Engine, a one-of-a-kind
machine designed by Charles Babbage in the 1820s, but famously never actually
completed, or at least not in his lifetime. The difference engine was a machine
designed to compute arbitrary functions, so it was a very early computer.

On the right, we have the inside of an Apple Watch S3 --- A solid state
computer, with no moving parts. Today, a company with over 120,000
employees, worth over one trillion dollars, makes and sells about 4 million of
these watches every quarter.

By just about every measure I can think of, the Apple Watch is a much better
computer than Babbage's 15 ton machine, and this is the kind of cumulative
cultural evolution I think we can all agree on.

Human populations solve increasingly complex problems over generations
that individuals struggle to solve on their own.

The Apple Watch solves a much more complex set of problems than the Difference
Engine. It has a touchscreen, sensors to detect your heartbeat, how many steps
you've taken, where you've been. There are engineering constraints on making it
wearable and sylish, and manufacturing constraints on how to produce
enough of these devices fast enough to keep up with demand.

Given this complexity, it's impossible to imagine any single person ever having
designed the Apple Watch, and this is an example of how cultures are better at
problem solving than individuals.
-->

# The psychology of cumulative cultural evolution

Why are cultures better at solving problems than individuals?

Inheritance
:   Knowledge is passed from one generation to the next.

Group size
:   Individuals can coordinate labor in large groups.

**Is it aggregation or interaction?**

<!--
Why are populations able to solve problems that individuals are not?

There are really two reasons I want to talk about today. The first is the idea
of inheritance: that cultures are able to pass knowledge from one generation to
the next, which means that problem solving can extend far longer than any
individual lifetime.

The second reason is related to group size. This is the idea that individuals
can coordinate their labor in groups to accomplish something that they couldn't
by themselves.

To me, as a psychologist, the most interesting question about both of these is
whether they represent a process of aggregation or interaction. Does problem
solving just add up as you add more generations or more people to
the group, or are there interactions, places where the group performs
differently than the sum of its parts.

I'm going to argue today that we actually know a fair amount about the effects
of group size on problem solving, but we actually don't know very much about
the effects on inheritance on problem solving.
-->

# When does the group outperform the individual?

Diverse groups can avoid premature convergence.[^diverse]  
Group decision making can improve through confidence matching.[^decision]  
Groups can sustain more complex innovations.[^groups]

[^diverse]: @Hong:2004gt
[^decision]: @Bahrami:2010jl
[^groups]: @Derex:2013kk


<!--
First let's look at what we know about where groups outperform the individuals
in them.

One of the more interesting findings in this line of research is where
diverse groups can outperform homogeneous`ones in cases where there
are locally optimal solutions. The diverse groups are able to maintain
a wider search area, and are less likely to converge prematurely.

A similar principle applies to diverse networks -- basically networks that are
not fully connected. In these cases, a sparsely connected network of 
individuals can outperform a fully connected network when the fully
connected network converges prematurely without fully searching the solution
space.

Another case where groups can outperform individuals is in some confidence
matching tasks, so in this study, by allowing dyads to confer their answers
in a perceptual decision making task, both individuals were able to perform
better than the best individual in the group responding alone.

Finally we have the cases where groups are able to sustain more complex
innovations simply through specialization, where individuals learn
to get good at one thing and learn from others rather than trying to
get good at everything at once.
-->

# Assembly bonus effects in group memory tasks

Groups remember more items, but each person's memory is not improved.

**Comparing groups to individuals requires adequate controls.**

# Measuring the effects of cultural inheritance on problem solving

```{r team-structures-1, out.height="80%", fig.align="center"}
crotchet::draw_graphviz("team-structures", package = "totems")
```

# Types of time

```{r types-of-time, fig.width=5, fig.height=2, fig.cap="Team structures controlling for labor time."}
gridExtra::grid.arrange(
  crotchet::read_graphviz("team-structures", package = "totems"),
  gg_time,
  gg_person,
  nrow = 1,
  widths = c(0.25, 0.375, 0.375)
)
```

# All team structure conditions

```{r team-structures-all, out.height="90%", fig.align="center"}
crotchet::draw_graphviz("team-structures-all", package = "totems")
```

# The Totem Game [@Derex:2015cbb]

```{r gameplay, out.height="90%"}
draw_image("gameplay", package = "totems")
```

# Recipes

```{r recipes, out.height="80%"}
crotchet::draw_image("recipes.gv", package = "totems")
```

# Solution landscape

```{r landscape, out.height="80%"}
crotchet::draw_image("landscape-sample.gv", package = "totems")
```

# Experiment 1: Simple diachronic inheritance

```{r team-structures-50min, fig.align="center"}
crotchet::draw_graphviz("team-structures-50min", package = "totems")
```

# Experiment 1: Simple diachronic inheritance

```{r innovations-50min}
grid.arrange(
  innovation_rate_50min_plot + ggtitle("A"),
  num_innovations_50min_plot + ggtitle("B"),
  nrow = 1,
  widths = c(0.55, 0.45)
)
```

# Experiment 2: Diachronic versus individual inheritance

```{r team-structures-selfother}
crotchet::draw_graphviz("team-structures-selfother", package = "totems")
```

# Experiment 2

```{r innovations-by-generation-selfother}
innovations_by_generation_plot
```

# Experiment 2

```{r diachronic-v-isolated-inheritance}
gridExtra::grid.arrange(
  learning_times_plot + ggtitle("A"),
  first_discovery_by_generation_plot + ggtitle("B"),
  ncol = 1
)
```

# Experiment 3

```{r scalability}
max_innovations_by_teamsize_plot
```

# Experiment 3

```{r scalability-simulations}
gridExtra::grid.arrange(
  max_innovations_by_teamsize_plot,
  bots_team_size_plot,
  nrow = 1
)
```

# Summary

Can cultural inheritance improve the problem solving abilities of future generations?

- A diachronic strategy is comparable when controlling for labor time.
- A diachronic strategy remediates the cost of fixation.
- A diachronic strategy is more scalable than a sychronic strategy.

# The effects of cultural inheritance on problem solving ability

**Pierce Edmiston**, Max Derex, Gary Lupyan  
pedmiston@wisc.edu  
[github.com/pedmiston/team-structures](https://github.com/pedmiston/team-structures)

# References {.allowframebreaks}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.25in}
\noindent
\footnotesize
