---
title: |
  The effects of cultural inheritance  
  on problem solving ability
author:
  \textbf{Pierce Edmiston}$^1$, Maxime Derex$^2$, Gary Lupyan$^1$
bibliography: references.bib
csl: apa.csl
output:
  beamer_presentation:
    theme: metropolis
    keep_tex: yes
    template: template/slides.tmpl
---
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(crotchet)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
# read_chunk("team-structures.R")
# read_chunk("iterated-problem-solving.R")
```

# Cumulative cultural evolution

```{r computer, fig.cap="Two computers. The Difference Engine (left) and the Apple Watch S3 (right)."}
gridExtra::grid.arrange(
  crotchet::read_image("img/difference-engine.jpg", width = 0.9),
  crotchet::read_image("img/apple-watch.jpg", width = 0.9),
  nrow = 1
)
```

<!--
Here are two computers.

The one on the left is a portion of the Difference Engine, designed by Charles
Babbage in the 1820s, but famously never actually built, or at least not in his
lifetime. The difference engine was a machine designed to compute arbitrary
functions, so it was a very early computer.

On the right, we have the inside of an Apple Watch S3 --- A solid state
computer, with no moving parts. Today, Apple, a company with over 120,000
employees, worth over one trillion dollars, makes and sells about 4 million of
these watches every few months.

I think we can all agree that the Apple Watch is a much better computer
Babbage's 15 ton machine, and this is the kind of cumulative cultural evolution
I think we can all agree on. The Apple Watch was not

The Apple Watch solves a much more complex set of
problems. It has a touchscreen, sensors to detect your heartbeat, how many
steps you've taken, where you've been. There are engineering constraints on
making it look wearable and sylish, and manufacturing constraints on how to
produce millions upon millions of these devices.

Babbage's machine and the Apple Watch are
separated by 145 years or about 8 generations. 


-->

# The psychology of cumulative cultural evolution

Groups solve increasingly complex problems over generations  
that individuals struggle to solve on their own.


# The two dimensions of cultural evolution

```{r dimensions-of-cultural-evolution, engine="dot"}
DiagrammeR::grViz('digraph {
  rankdir=LR;
  node[label="" shape=circle];
  a1 -> a2[minlen=4];
  b1, b2, b3, b4;
  b2, b3[style=invis];
  b1 -> b4[constraint=false];
}')
```




Where I think we may disagree is in what I will call the psychology
of cumulative cultural evolution.
-->

# The psychology of cumulative cultural evolution

> Humans are adapted to improve one another's inventions.

- Human tools have improved over generations faster than tools used by other animal species.



```{r iterated-problem-solving, engine="dot"}
digraph {
    rankdir = LR
    node[shape=circle style=filled label=""]
    bgcolor="transparent"

    d1 -> d2 -> d3 -> d4 -> d5
    d5[style=invis]

    d1[fillcolor="#7483A7"];
    d2[fillcolor="#8DA0CB"];
    d3[fillcolor="#A1B1D4"];
    d4[fillcolor="#B6C2DD"];
}
```

# The adaptiveness of cultural inheritance[^br]

Cultural inheritance is adaptive because **even smart people die.**  
Cultural inheritance enables problem solving to extend far longer than any lifetime.

[^br]: [@Boyd:1985cu; @Richerson:2005not]

```{r adaptiveness, engine="dot"}
digraph {
    rankdir = LR
    node[shape=circle style=filled label=""]
    bgcolor="transparent"

    d1, i2  // set row order last to first

    d1 -> d2 -> d3 -> d4 -> d5
    i1 -> i2
    d5, i2[style=invis]

    i1[fillcolor="#66C2A5"]

    d1[fillcolor="#7483A7"];
    d2[fillcolor="#8DA0CB"];
    d3[fillcolor="#A1B1D4"];
    d4[fillcolor="#B6C2DD"];
}
```

# Increasing acquisition costs[^var]

```{r mesoudi-2011-fig2a}
crotchet::draw_image("img/mesoudi-2011-fig2a.png")
```

[^var]: [Fig. 2A, @Mesoudi:2011fn]

# The Totem Game [@Derex:2015cbb]

```{r gameplay, out.height="90%"}
draw_image("gameplay", package = "totems")
```

<!--
An example of this is the Totem game. In the Totem game, participants attempt
to build a sacred totem to appease the gods. To build a totem pole, first they
have to combine these resources to make tools, and use
those tools to create other tools, and so on and so on, until they can chop
down a tree, and carve and paint a totem pole.

In this game, most of the combinations participants try do not yield new tools,
and so the test is how effectively participants can search the space of possible
combinations without making the same incorrect guesses over and over again.
-->

# Recipes

```{r recipes, out.height="80%"}
crotchet::draw_image("recipes.gv", package = "totems")
```

<!--
Here are some of the successful combinations, which I will refer to
as "recipes". You can see a few different kinds of recipes here,
some where the same item is combined with itself, others where
you can refine an item all by itself, but most of the recipes
involve the combination of 2 to 3 different items.

What's important about these recipes is that they are what
is inherited by future generations. First generation participants
had to discover all these recipes on their own, but other
participants started off with access to the recipes discovered
by the previous generation. So they could click on an item,
and see its recipe, so that they could recreate it on their own.
-->

# Solution landscape

```{r landscape, out.height="80%"}
crotchet::draw_image("landscape-sample.gv", package = "totems")
```

<!--
What makes this game interesting is that as tools are accumulated, the space
of possible combinations increases exponentially, even though the space
of correct combinations doesn't.

Here I'm showing the first 6 generations of tools, leading to the first axe,
which is required to make the simplest kind of totem pole. The landscape goes
on beyond this point, but in reality, in just 25 minutes, participants have
a hard time getting even this far.

Presumably, with more items, the game gets harder and harder, so we can ask
how individuals who inherit a bunch of items from an ancestor are able to
exceed them. Are they able to overcome the acquisition cost of recreating
the previous tools?
-->

# References {.allowframebreaks}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.25in}
\noindent
\footnotesize
