### Simulations {-}

We used an agent-based simulation to model the scalability of diachronic and
synchronic strategies. In the simulations, problem solvers guessed combinations
of between 1 and 4 items at random. Diachronic and synchronic strategies were
provided the same number of guesses. Diachronic problem solvers guessed one at
a time, while synchronic problem solvers guessed in parallel. If a guess
yielded an item, it was added to the team inventory. For synchronic problem
solvers, guesses were added to each participant's inventory immediately.

The results show that as problem solvers are added to each team, a synchronic
strategy becomes progressively less effective as compared to a diachronic
strategy (Fig. \@ref(fig:bots-teamsize)). The reason is that synchronic players
increasingly make the same incorrect guesses because each player is generating
guesses from the same inventory. Diachronic players in contrast are not as
likely to generate the same incorrect guesses as a previous generation because
they are generating guesses from different sets of items.

(ref:bots-teamsize) Result of simulations of diachronic and synchronic problem solvers.

```{r bots-teamsize, fig.width=5, fig.cap="(ref:bots-teamsize)"}
bots_team_size_plot
```

### Guessing strategy {-}

Although participants in the isolated strategy discovered more tools
after four sessions than diachronic teams after four generations,
they did not engage in a more efficient guessing strategy. As in Experiment 1,
participants in the diachronic condition had the most effective guessing
strategy in that they generated a lower proportion
of redundant guesses than participants in the isolated condition,
but in Experiment 2, this different was not statistically significant,
`r exp2$prop_redundant_strategy`. For participants in both strategies,
the proportion of redundant guesses decreased over generations and sessions,
`r exp2$prop_redundant_gen`. However, this decrease was not affected
by strategy, `r exp2$prop_redundant_inter`.

For the proportion of unique guesses, there were no differences between
participants in either strategy, `r exp2$prop_unique_strategy`, or
over generations `r exp2$prop_unique_gen`. The change in proportion
of unique guesses over generations was not affected by strategy,
`r exp2$prop_unique_inter`.

(ref:guessing-strategy-by-generation) Proportion of guesses by participants in
diachronic and isolated conditions over generations.

```{r guessing-strategy-by-generation, fig.width=5, fig.cap="(ref:guessing-strategy-by-generation)"}
prop_guess_types_selfother_plot
```

If participants in the isolated strategy are not guessing more efficiently
than participants in the diachronic strategy, why then are they able
to accumulate more tools than participants in the diachronic condition?
To answer this question, we investigated the time spent recreating
the inherited tools versus time spent inventing new tools.